prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-30 inst/MyArticle/paper.Rmd
edits

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008] which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of quantitative proteomics data. However, the existing packages differ widely regarding supported designs, model specifications, and output formats. At the same time, they have the following features in common: fitting linear models to either peptide or protein intensities, determining differences among groups, and afterward applying empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented design, with R linear models as a core, and add features such as $p$-value aggregation, variance shrinkage, or modeling of missing observations.

Furthermore, `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Peptide or protein abundances can then be scaled and transformed using robust scaling, _quantile_ normalization or `r Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with R packages such as `r Biocpkg("DEP")` [@DEP2018] or `r Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$). We use the IonStar [@shen2018ionstar] and CPTAC[@edwards2015cptac]  datasets, processed with *MaxQuant* and *FragPipe* to benchmark the modeling methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r Biocpkg("MSstats")`, `r Biocpkg("msqrob2")` and `r Biocpkg("proDA")`.

Although other benchmark datasets exist[@wessels2012comprehensive;@o2018proteome], the IonStar dataset has the advantage that the expected differences of the spike in proteins among groups are slight compared to other benchmark datasets, making DEA more difficult and enabling us to see performance differences among the methods. For the same reason of making the DEA more challenging and enhancing the differences between the statistical modeling approaches, we focus on the MaxQuant outputs since they are noisier than those produced by FragPipe.

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an MSFragger[@yu2020fast] configuration can be found in the Appendix. We implemented methods that transform the data into tidy data for popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features (see Supplementary Figure SX) in RStudio [@RStudio2022], and help to implement argument free functions.

@@

R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r Biocpkg("proDA")` or of *SAINTexpress* [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

```{r prolfquaModels}
xM <- data.frame( 
    lm = c("strategy_lm, Contrasts", "linear modelling of peptide or protein abundances and group difference estimation"),
    lmer = c("strategy_lmer, Contrasts", "mixed effect modelling of peptide or protein abundances and group differences estimation" ),
    gmi = c("ContrastsSimpleImpute" , "group difference estimation when no observations in one of the groups" ),
    ROPECA = c("ContrastsROPECA", "estimating group differences for proteins by summarizing peptide differences"),
    limma = c( "ContrastsModerated", "empirical Bayes variance shrinkage for group difference estimates (limma)"),
    pic = c("runSaint, ContrastsSaintExpress", "protein interaction scoring (SAINTExpress)"),
    prodA = c( "strategy_proDA*, ContrastsProDA", "adapter to the probabilistic dropout model implemented in proDA"), check.names =  FALSE)
xM <- data.frame(t(xM))

@@

To evaluate the performance of differential expression analysis, we use the IonStar benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells (Panc-1) lysate at five different levels: $3\%$, $4.5\%$, $6\%$, $7.5\%$ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters _a_, _b_, _c_, _d_ and _e_. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, hence 20 raw files in total. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

However, other  have similar constraints. Furthermore, we could not analyze the benchmark dataset [@wessels2012comprehensive] with a complete factorial design with an up-to-date version of MaxQuant.

@@

We processed the raw data of the IonStar dataset using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings for orbitrap data. The text files generated by MaxQuant are available in the *prolfquadata* R package [@prolfquadata]. MaxQuant produces various output files which can be used for DEA. We are using the quantification results reported in the 'peptide.txt' file for DEA. However, *MSstats* is using the 'evidence.txt' file for the DEA.

@@

We processed the raw data of the IonStar dataset using the FragPipe [@yu2020fast] Version 14, with default workflow for label free quantification with match between runs. The text files generated by FragPipe are available in the *prolfquadata* R package [@prolfquadata].
Similarily to MaxQuant, the FragPipe produces various outputs which can e used for DEA. We used the total protein intensities reported in the `combined_protein.tsv` file as input for the DEA and call this dataset _IonStar/FragPipe/combined_protein.tsv_ Alternatively, we benchmarked the DEA using the MSstats.tsv file as input and call this dataset _IonStar/FragPipe/combined_protein.tsv_.

@@

The peptide abundances (from the *MaxQuant* _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean absolute deviation were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r Biocpkg("proDA")` to protein abundances, the mixed effect linear model, the `r Biocpkg("PECA")` model, and hurdle model `r Biocpkg("msqrob2")` to peptide intensities.

@@

The IonStar dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins, the difference $\beta$ between two dilutions should be $\beta = 0$, while for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$.

@@

The package `r BiocStyle::Githubpkg("fgcz/prolfqua")` provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string. In the Supplement we provided an example, how to specify contrasts for a dataset with two explanatory variable with an interaction term (Supplementary Section "Specifying Contrasts for Models with two Factors and Interaction Term").

@@

However, if there is no observation in a group the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample (see Supplement section Estimating $A_{LOD}$). Typically there are many such cases, and hence we take the median.

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which *MaxQuant* reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:figure1) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:figure1) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

R linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. Models in `r BiocStyle::Githubpkg("fgcz/prolfqua")` are specified using R's linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending; @MASS2002] is advantageous when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` using base R or other programming languages by reading the analysis script.
However, in the package documentation, we showcase how a user, without this knowledge, can analyze experiments with a parallel group design and a factorial design.

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`'s functionality (see Supplementary Figure S[X]).

To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the prolfqua package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric;@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula, and which comparisons to compute from the sample annotation provided in tabular form, similarly to *SAINTexpress* or *MSstats*. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, `r BiocStyle::Githubpkg("fgcz/prolfqua")` and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and helps scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR  [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with the results produced by *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4178 proteins with at least two peptides, there are in total $16712$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that *MSstats* estimates $16038$ group differences while the mixed effect models estimates the fewest with $15756$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

@@

We obtained difference and FDR estimates for all proteins, as shown in Figure \@ref(fig:figure3) Panel A when using: a) the probabilistic dropout model (*proDA*), b) the hurdle model (*msqrob2*), and c) `prolfqua_merge` (Table \@ref{tab:describeBenchmarkedModels}). We observe that the performance of the scaled $p$-values or the $t$-statistics is comparable among these three methods (Figure \@ref(fig:figure3) Panel B). We tested if there is a significant difference between the $pAUC_{10}$ for all three methods, but did not reject the null hypothesis that there is no such difference (Supplementary Section "Comparing pAUC for IonStar/MaxQuant dataset"). Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods. Furthermore, all three models perform similarly well when examined using CPTAC/MaxQuant dataset (see Supplementary section "Benchmarking msqrob2, proDA and prolfqua using the CPTAC dataset"). In addition we also examined the performance of the *proDA* and `prolfqua_merged` model using the IonStar dataset but processed with the `FragPipe v14` software. While, for the IonStar/MaxQuant dataset, *msqrob2* performed best, *proDA* performed best for the CPTAC/MaxQuant data, and `prolfqua_merged` for the IonStar/FragPipe dataset.

We focused our benchmark on comparing the statistical modeling methods while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supplement we show the benchmarking results for *proDA* and *prolfqua* when the IonStar dataset is prepossessed with the FragPipe v14 [@yu2020fast] quantification pipeline. Although the number of proteins identified with two peptides is slightly lower with FragPipe than with MaxQuant (3929 vs. 4046), the $pAUC_{10}$ is significantly higher ($81%$ vs. $72%$).

>>>>

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008] which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the DEA of quantitative proteomics data. However, the existing packages differ widely regarding supported designs, model specifications, and output formats. At the same time, they have the following features in common: fitting linear models to either peptide or protein intensities, determining differences among groups, and afterward applying empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented design, with R linear models as a core, and add features such as $p$-value aggregation, variance shrinkage, or modeling of missing observations.

Furthermore, `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Peptide or protein abundances can then be scaled and transformed using robust scaling, _quantile_ normalization or `r Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, *prolfqua* can be compared with R packages such as `r Biocpkg("DEP")` [@DEP2018] or `r Biocpkg("POMA")` [@POMA2021] which support the entire 
DEA pipeline.

Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$). We use the *IonStar*[@shen2018ionstar] and *CPTAC*[@edwards2015cptac]  datasets, processed with *MaxQuant* and *FragPipe* to benchmark the modeling methods implemented in *prolfqua* and to compare our results with those of *MSstats*, *msqrob2* and *proDA*.

Although other benchmark datasets exist[@wessels2012comprehensive;@o2018proteome], the *IonStar* dataset has the advantage that the expected differences among groups of the spike in proteins are slight compared to other benchmark datasets, making DEA more difficult and enabling us to see performance differences. For the same reason, of making the DEA more challenging and enhancing the differences between the statistical modeling approaches, we focus on the *MaxQuant* outputs since they are noisier than those produced by *FragPipe*.

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in *prolfqua* if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in *MSstats*[@Choi2014] format, only a few lines of code are needed to update the *prolfqua* `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an *FragPipe*[@yu2020fast] configuration can be found in the Supplementary Section: "Creating a prolfqua configuration". We implemented methods that transform the data into tidy data for popular software like *MaxQuant*[@Cox2008MaxQuant], or *FragPipe*, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features (see Supplementary Figure 8) in *RStudio* [@RStudio2022], and help to implement argument free functions.

@@

R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with *prolfqua* in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in *prolfqua* (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform DEA given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate DEA results of external tools such as *proDA* or of *SAINTexpress* [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

```{r prolfquaModels}
xM <- data.frame( 
    lm = c("strategy_lm, Contrasts", "linear modelling of peptide or protein abundances and group difference estimation"),
    lmer = c("strategy_lmer, Contrasts", "mixed effect modelling of peptide or protein abundances and group differences estimation" ),
    gmi = c("ContrastsSimpleImpute" , "group difference estimation when no observations in one of the groups" ),
    ROPECA = c("ContrastsROPECA", "estimating group differences for proteins by summarizing peptide differences"),
    limma = c( "ContrastsModerated", "empirical Bayes variance shrinkage for group difference estimates (limma)"),
    pic = c("runSaint, ContrastsSaintExpress", "protein interaction scoring (SAINTExpress)"),
    proDA = c( "strategy_proDA*, ContrastsProDA", "adapter to the probabilistic dropout model implemented in proDA"), check.names =  FALSE)
xM <- data.frame(t(xM))

@@

To evaluate the performance of DEA, we use the *IonStar* benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells (Panc-1) lysate at five different levels: $3\%$, $4.5\%$, $6\%$, $7.5\%$ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters _a_, _b_, _c_, _d_ and _e_. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, hence 20 raw files in total. To compare the performance of the various methods implemented in *prolfqua* we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

@@

We processed the raw data of the *IonStar* dataset using *MaxQuant* [@Cox2008MaxQuant] version Version 1.6.10.43, with *MaxQuant* default settings for orbitrap data. The text files generated by *MaxQuant* are available in the *prolfquadata* R package [@prolfquadata]. *MaxQuant* produces various output files which can be used for DEA. We are using the quantification results reported in the 'peptide.txt' file for DEA. However, *MSstats* is using the 'evidence.txt' file for the DEA.

@@

We processed the raw data of the *IonStar* dataset using the *FragPipe* [@yu2020fast] Version 14, with the default workflow for label free quantification with match between runs. The text files generated by *FragPipe* are available in the *prolfquadata* R package [@prolfquadata].
Similarily to *MaxQuant*, the *FragPipe* software produces various outputs which can be used for DEA. We used the total protein intensities reported in the 'combined_protein.tsv' file as input for the DEA and call this dataset IonStar/FragPipe/combined_protein.tsv. Alternatively, we benchmarked the DEA using the 'MSstats.tsv' file as input and call this dataset IonStar/FragPipe/combined_protein.tsv.

@@

The peptide abundances (from the *MaxQuant* _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean absolute deviation were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model *proDA* to protein abundances, the mixed effect linear model, the *ROPECA* model, and hurdle model implemented in *msqrob2* to peptide intensities.

@@

The *IonStar* dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins, the difference $\beta$ between two dilutions should be $\beta = 0$, while for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$.

@@

The package *prolfqua* provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string. In the Supplement we provided an example, how to specify contrasts for a dataset with two explanatory variable with an interaction term (Supplementary Section "Specifying Contrasts for Models with two Factors and Interaction Term").

@@

However, if there is no observation in a group the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample (see Supplementary Section "Estimating $A_{LOD}$"). Typically there are many such cases, and hence we take the median.

@@

The code snippets in this section demonstrate how a DEA workflow can be implemented using the *prolfqua* R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which *MaxQuant* reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:figure1) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:figure1) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

R linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. Models in *prolfqua* are specified using R's linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending; @MASS2002] is advantageous when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with *prolfqua* using base R or other programming languages by reading the analysis script.
However, in the package documentation, we showcase how a user, without this knowledge, can analyze experiments with a parallel group design and a factorial design.

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of *prolfqua* with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the *RStudio* editor makes it easy for novices to explore *prolfqua*'s functionality (see Supplementary Figure 8).

To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the *prolfqua* package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric;@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula, and which comparisons to compute from the sample annotation provided in tabular form, similarly to *SAINTexpress* or *MSstats*. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, *prolfqua* and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and help scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR  [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of *prolfqua* includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in *prolfqua* model quantitative mass spectrometric high throughput data and compare them with the results produced by *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

When comparing DEA performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4178 proteins with at least two peptides, there are in total $16712$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that *MSstats* estimates $16038$ group differences while the mixed effect models estimates the fewest with $15756$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

@@

We obtained difference and FDR estimates for all proteins, as shown in Figure \@ref(fig:figure3) Panel A when using: a) the probabilistic dropout model (*proDA*), b) the hurdle model (*msqrob2*), and c) `prolfqua_merge` (Table \@ref{tab:describeBenchmarkedModels}). We observe that the performance of the scaled $p$-values or the $t$-statistics is comparable among these three methods (Figure \@ref(fig:figure3) Panel B). We tested if there is a significant difference between the $pAUC_{10}$ for all three methods, but did not reject the null hypothesis that there is no such difference (Supplementary Section "DEA benchmark : IonStar/MaxQuant/peptide.txt - Significance test"). Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods.

Furthermore, all three models perform similarly when examined using a different benchmark dataset CPTAC/MaxQuant (see Methods). For this dataset, *proDA* performed slightly but not significantly better than *prolfqua* and *msqrob* (and Supplementary Section "DEA benchmark : CPTAC/MaxQuant/peptide.txt").

In addition, we examined the DEA performance when using protein intensities reported by quantification software *FragPipe* for the *IonStar* dataset as input. Using protein abundances as input significantly simplifies the analysis and interpretation and might benefit from optimization implemented in the signal processing software. However, we can only easily fit the *proDA* and `prolfqua_merged` models to protein intensities. In this DEA benchmark, *prolfqua* performed slightly, but not significantly better than *proDA*  (Supplementary Section "DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv").

Finally, we also compared the DEA performances when starting the analysis from the precursor abundances reported in the 'MSstats.tsv' file, generated by *FragPipe* v14, from the *IonStar* dataset. Since *MSstats*, *msqrob2*, *proDA*, and *prolfqua* all read `MStstats.tsv` files, we could eliminate a confounding factor, i.e., different input abundances (Supplementary Section "DEA benchmark: IonStar/FragPipeV14/MSstats.tsv"). In this DEA benchmark, *msqrob2* and *prolfqua* perform best but not significantly better than *proDA* or *MSstats*. Furthermore, by processing the *IonStar* dataset with both *MaxQuant* and *FragPipe*, we can compare their performances (Supplementary Section "Comparing DEA results for *MaxQuant* and *FragPipe*").

We focused our benchmark on comparing the statistical modeling methods while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by *MaxQuant*, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supplementary Section "Comparing DEA results for *MaxQuant* and *FragPipe* we compare DEA benchmarking results for *MaxQuant* and *FragPipe*. While the number of proteins identified with two peptides is practically the same, the DEA benchmark performance differs significantly by $\approx 10%$ for the $pAUC_{10}$ score.
